---
title: "Week 5 assignment -- example solution"
author: "Pierre-Luc"
date: "4/1/2022"
output: html_document
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE) 
```

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(epiwraps)
})
```

# Getting the data

## Download 

```{r, eval=FALSE}
dir.create("peaks")
dir.create("tracks")
options(timeout=1800)

# mESC
download.file("https://www.encodeproject.org/files/ENCFF008XKX/@@download/ENCFF008XKX.bed.gz", "peaks/H3K27me3_mESC.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF160FEV/@@download/ENCFF160FEV.bigWig", "tracks/H3K27me3_mESC.bw")

download.file("https://www.encodeproject.org/files/ENCFF247GVM/@@download/ENCFF247GVM.bed.gz", "peaks/H3K4me3_mESC.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF611GSQ/@@download/ENCFF611GSQ.bigWig", "tracks/H3K4me3_mESC.bw")

# myocytes
download.file("https://www.encodeproject.org/files/ENCFF044YOI/@@download/ENCFF044YOI.bed.gz", "peaks/H3K27me3_myocyte.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF045JOU/@@download/ENCFF045JOU.bigWig", "tracks/H3K27me3_myocyte.bw")

download.file("https://www.encodeproject.org/files/ENCFF944EDY/@@download/ENCFF944EDY.bed.gz", "peaks/H3K4me3_myocyte.bed.gz")
download.file("https://www.encodeproject.org/files/ENCFF503YQC/@@download/ENCFF503YQC.bigWig", "tracks/H3K4me3_myocyte.bw")
```

## Reading the peaks into R

```{r}
peakfiles <- list.files("peaks/", full=TRUE)
names(peakfiles) <- gsub(".bed.gz", "", basename(peakfiles), fixed=TRUE)
sets <- lapply(peakfiles, FUN=function(x) rtracklayer::import(x, format="narrowPeak"))
```

# Identifying bivalent domains

```{r}
# method one: find the H3K4me3 peaks that overlap a H3K27me3 domain:
biv1 <- sets$H3K4me3_mESC[overlapsAny(sets$H3K4me3_mESC, sets$H3K27me3_mESC)]
# method two: fine the regions that are covered by both H3K4me3 and H3K27me3
biv2 <-intersect(sets$H3K4me3_mESC, sets$H3K27me3_mESC)

table(overlapsAny(biv1, biv2))
table(overlapsAny(biv2, biv1))
```

The number and sizes of the regions are different, but all regions of one set overlap some regions of the other.

# Answering the questions

How many of the mESC bivalent domains have either mark or their combination in myocytes?

```{r}
isK27 <- overlapsAny(biv2, sets$H3K27me3_myocyte) & !overlapsAny(biv2, sets$H3K4me3_myocyte)
table(isK27)

isK4 <- overlapsAny(biv2, sets$H3K4me3_myocyte) & !overlapsAny(biv2, sets$H3K27me3_myocyte)
isBiv <- overlapsAny(biv2, sets$H3K4me3_myocyte) & overlapsAny(biv2, sets$H3K27me3_myocyte)
```

Of the `r length(isK27)` mESC bivalent domains, `r sum(isK27)` are repressed (H3K27me3 without H3K4me3) in myocytes.
Of the `r length(isK4)` mESC bivalent domains, `r sum(isK4)` are activated (H3K4me3 without H3K27me3) in myocytes.
Of the `r length(isBiv)` mESC bivalent domains, `r sum(isBiv)` remain bivalent in myocytes.


Or we could use the upset plot:

```{r}
sets2 <- list(bivalent_mESC=biv2, H3K4me3_myocyte=sets$H3K4me3_myocyte,
              H3K27me3_myocyte=sets$H3K27me3_myocyte)
regionUpset(sets2)
regionUpset(sets2, reference = biv2)
```

What we can observe is that the vast majority of mESC bivalent domains (4058 out of 4294) become either repressed or activated upon differentiation (with more becoming repressed).

